Equalizer 2.0 --\\Convergence of a Parallel Rendering Framework

### 2017-12-14 paperIEEE.tex
Review Usability section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@71 db7c848b-7755-4df2-b763-8796797afa52

Real-world visualization setups are often complex, and having an abstract
representation of the display system can simplify the configuration
process. Real-world applications often have the need to be aware of spatial
relationship of the display setup, for example to render 2D overlays or to
configure multiple views on a tiled display wall.

@@

A typical installation consists of one projection canvas, which is one
aggregated projection surface, e.g., a tiled display wall or a CAVE. Desktop
windows are considered a canvas. Each canvas is made of one or more segments,
which are the individual outputs connected to a display or projector. Segments
can be planar or non-planar to each other, and can overlap or have gaps between
each other. A segment is referencing a channel, which defines the output area of
this segment, e.g., on a DVI connector connected to a projector.

A canvas can define a frustum, which will create default, planar sub-frusta for
all of its segments. A segment can also define a frustum, which overrides the
canvas frustum, e.g., for non-planar setups such as CAVEs or curved
screens. These frusta describe a physically-correct display setup for a Virtual
Reality installation. \label{sec:swap} A canvas may have a software or hardware
swap barrier, which will synchronize the rendering of all contributing GPUs. The
software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all
segments, the
swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
the swap command. The hardware swap barrier is implemented using the
\textsf{WGL\_NV\_swap\_group} or \textsf{GLX\_NV\_swap\_group} extensions.

On each canvas, the application can display one or more views. A view is a view
on a model, in the sense used by the MVC pattern. The view class is used by
\textsf{Equalizer} applications to define view-specific data for rendering,
e.g., a
scene, viewing mode or camera. The application process manages this data, and
the render clients receive it for rendering.

@@

A view may have a frustum description. The view's frustum overrides frusta
specified at the canvas or segment level. This is used for non-physically
correct rendering, e.g., to compare two models side-by-side on a tiled display
wall. If the view does not specify a frustum, the corresponding destination
channels will use the physically correct sub-frustum resulting from the
view/segment intersection.

\label{sec:observer}
An observer looks at one or more views. It is described by the observer position
in the world and its eye separation. Each observer will have its own stereo
mode, focus distance and frame loop (framerate). This allows to have untracked
views and multiple tracked views, e.g., two HMDs, in the same application.

@@

Switching a layout, as described above, or switching the stereo rendering mode,
may involve a different set of resources after the change, including the launch
and exit of render client processes. \textsf{Equalizer} solves this through a
reconfiguration step at the beginning of each rendering frame. Each resource
(channel, window, pipe, node) has an activation count, which is updated when the
layout or any other relevant rendering parameter is changed. When a resource is
found whose activation count does not match its current start/stopped state, the
resource is created or destroyed and \textsf{configInit} or \textsf{configExit}
are called accordingly. In the current implementation, a normal configuration
initialization or exit, as described in \cite{EMP:09}, uses the same code path
with all used resources transitioning to a running or stopped state,
accordingly. Since starting new resources typically requires object mapping and
associated data distribution, it is a costly operation. To improve performance
and robustness, \textsf{Equalizer} will flush the rendering pipeline whenever a
runtime
reconfiguration is detected, before executing the change. This implementation
allows the late join or early exit, transparent to application developers.

@@

This versatile mechanism allows non-experts to configure and profit from
multi-GPU workstations and visualization clusters, as well as to provide system
administrators with the tools to implement easy to use integration with cluster
schedulers. This feature is transparent to \textsf{Equalizer} application
developers.

@@

Qt is a popular window system with many application developers. Unfortunately,
it imposes a different threading model for window creation and event handling
compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering
thread is independently
responsible for creating its windows, receiving the events and eventually
dispatching them to the application process main thread. This design is
motivated by the natural threading model of X11 and WGL, and allows simple
sequential semantics between OpenGL rendering and event handling. In contrast,
Qt requires all windows and QOpenGLContext to be created from the Qt main
thread. An existing Qt window or context may be moved to a different thread, and
events are signalled from the main thread.

Challenges in threading model, architecture

@@

\textsf{Sequel} is a simplification layer for \textsf{Equalizer}. It is based
on the realization
that while fully expressive, the verbatim abstraction layer of nodes, pipes,
windows and channels in \textsf{Equalizer} requires significant learning to
fully
understand and exploit. In reality, a higher abstraction of \textsf{Application}
and \textsf{Renderer} is sufficient for most use cases. In \textsf{Sequel}, the
application class drives the configuration, and one renderer instance is created
for each (pipe) render thread. They also provide the natural place to store and
distribute data. Last, but not least, \textsf{ViewData} provides a convenient
way to manage multiple views by storing the camera, model or any other
view-specific information.

>>>>

Real-world visualization setups can be complex. An abstract representation of
the display system simplifies the configuration process. Applications often have
the need to be aware of spatial relationship of the display setup, for example
to render 2D overlays or to configure multiple views on a tiled display wall.

@@

A typical installation consists of one projection canvas, which is one
aggregated projection surface, e.g., a tiled display wall or a CAVE. Desktop
windows are considered a canvas. Each canvas is made of one or more segments,
which are the individual outputs connected to a display or projector. Segments
can be planar or non-planar to each other, and can overlap or have gaps between
each other. A segment is referencing a channel, which defines the output area of
this segment, e.g., on a DVI connector connected to a projector. This
abstraction covers all use cases from simple windows, tiled display walls with
bezels, to non-planar immersive systems with edge-blending.

A canvas can define a frustum, which will create default, planar sub-frusta for
all of its segments. A segment can also define a frustum, which overrides the
canvas frustum, e.g., for non-planar setups such as CAVEs or curved screens.
These frusta describe a physically-correct display setup for a Virtual Reality
installation. \label{sec:swap} A canvas may have a software or hardware swap
barrier, which will synchronize the rendering of all contributing GPUs. The
software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all
segments, the swap buffers call followed by a \textsf{glFlush} to ensure timely
execution of the swap command. The hardware swap barrier is implemented using a
\textsf{NV\_swap\_group} extension.

On each canvas, the application can display one or more views. A view is used in
the sense of the MVC pattern. The view class is available to applications to
define view-specific data for rendering, e.g., a scene, viewing mode or camera.
The application process manages this data, and the render clients receive it for
rendering.

@@

% A view may have a frustum description. The view's frustum overrides frusta
% specified at the canvas or segment level. This is used for non-physically
% correct rendering, e.g., to compare two models side-by-side on a tiled display
% wall. If the view does not specify a frustum, the corresponding destination
% channels will use the physically correct sub-frustum resulting from the
% view/segment intersection.

\label{sec:observer}
An observer looks at one or more views. It is described by the observer position
in the world and its eye separation. Each observer has its own stereo
mode, focus distance and eye positions. This allows to have untracked
views and multiple tracked views, e.g., two HMDs, in the same application.

@@

Switching a layout, as described above, or switching the stereo rendering mode,
may involve a different set of resources after the change, including the launch
and exit of render client processes. \textsf{Equalizer} solves this through a
reconfiguration step at the beginning of each rendering frame. Each resource
(channel, window, pipe, node) has an activation count, which is updated when the
layout or any other relevant rendering parameter is changed. When a resource is
found whose activation count does not match its current start/stopped state, the
resource is created or destroyed and \textsf{configInit} or \textsf{configExit}
are called accordingly. In the current implementation, a normal configuration
initialization or exit, as described in \cite{EMP:09}, uses the same code path
with all used resources transitioning to a running or stopped state,
accordingly. Since starting new resources typically requires object mapping and
associated data distribution, it is a costly operation.

@@

This versatile mechanism allows non-experts to configure and profit from
multi-GPU workstations and visualization clusters, as well as to provide system
administrators with the tools to implement easy to use integration with cluster
schedulers. This feature is transparent to application developers.

@@

Qt is a popular window system with many application developers. Unfortunately,
it imposes a different threading model for window creation and event handling
compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread
is independently responsible for creating its windows, receiving the events and
eventually dispatching them to the application process main thread. This design
is motivated by the natural threading model of X11 and WGL, and allows simple
sequential semantics between OpenGL rendering and event handling. In contrast,
Qt requires all windows and QOpenGLContext to be created from the Qt main
thread. An existing Qt window or context may subsequently be moved to a
different thread, and events are signalled from the main thread. For Qt windows,
Equalizer will internally dispatch and handle the window creation from the
render to the main thread, move the created objects back to the render thread,
and dispatch Qt signals to the correct render threads.

@@

\textsf{Sequel} is a simplification layer for \textsf{Equalizer}. It is based on
the realization that while fully expressive, the verbatim abstraction layer of
nodes, pipes, windows and channels in \textsf{Equalizer} requires significant
learning to fully understand and exploit. In reality, a higher abstraction of
only \textsf{Application} and \textsf{Renderer} is sufficient for many use
cases. In \textsf{Sequel}, the application class drives the configuration, and
one renderer instance is created for each (pipe) render thread. They also
provide the natural place to store and distribute data. Last, but not least,
\textsf{ViewData} provides a convenient way to manage multiple views by storing
the camera, model or any other view-specific information.
